Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2021-07-14 paper/paper.tex
WIP on paper now I know all interactions are receptor-ligand based

\emph{Abstract here.}

@@

Our model was inspired by the agent-based model of \cite{simpson_simple_2011}
and, in common with that work, consists of agents representing axonal growth
cones, originating from a square retina and moving upon a square tectum. Also
in common with the Simpson \& Goodhill model, we model three contributions to
the movement of the growth cones; chemoaffinity, competition and axon-axon
interaction. In our model, chemoaffinity and axon-axon interactions occur via
signal transmission between receptors and ligands which are expressed in
graded patterns on both retina and tectum. This sets our model apart from that
of Simpson \& Goodhill, in which each growth cone moving on the tectum was
always able to determine a vector to its designated target and advance along
that vector. Competition is modelled as a distance based repulsion as
in \cite{simpson_simple_2011}.

@@

Because there is convincing evidence that EphA and EphB receptors are
expressed in exponentially increasing
patterns \cite{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004},
we use an exponential form for retinal receptor expressions, in common with
other modelling
studies~\cite{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
%
The tectum expresses the same sets of receptors and ligands, also in
orthogonal pairs of gradients. In this work, we did not consider the effect of
reverse signalling from tectal receptors to RGC ligands so we only modelled
tectal ligand expression. Although several studies model tectal ligand
expression with exponential functions~\cite{koulakov_stochastic_2004}, the
experimental evidence for ligand expression is more ambiguous (Supp. Fig
X). Correspondingly, we leave open the possibility that ligand expression may
be modelled by exponential, linear or logarithmic functions.

While we do not explicitly name $r_0$, $r_1$,
etc.~as \emph{EphA}, \emph{EphB}, the suggestion is that $r$ includes EphA,
EphB, Ryk~\cite{schmitt_wntryk_2006} and
Neogenin~\cite{rajagopalan_neogenin_2004} receptors and that $l$ includes the
ephrin-A, ephrin-B, Wnt3~\cite{schmitt_wntryk_2006} and
RGM~\cite{monnier_rgm_2002} ligands, each of which has been shown to play a
role in retino-tectal map formation.

@@

In this model, the signal transmitted when a ligand binds to a receptor on a
growth cone can lead to one of two effects; the signal may induce the growth
cone to climb the gradient of ligand expression, in which case the interaction
is \emph{attractive} or the cone may descend the gradient of ligand expression
if the interaction is \emph{repulsive}. To determine the strength of the
effect we could assume a purely linear receptor binding model, and set the
movement of the growth cone to be $\sum_i \pm r_i \times \grad l_i$. While
this approach works, it does not allow us to investigate the effect of
receptor binding noise (\textbf{FIXME}: Be sure to have included a discussion
on the three types of noise as described by Goodhill, 2016 in the
introduction) on the ability of the system to generate topographic order.

Instead, we treat the receptor binding at the growth cone with greater
care. Based on a Michaelis-Menten model applied to
receptor binding, \citet{mortimer_bayesian_2009} give an expression for the
probability of an individual receptor being bound as a function of the ligand
expression, $\gamma$, and the ligand gradient, $\mu$;
\begin{equation}
P(b=1|\gamma,\mu) = \frac{\gamma(1 + \mu r)}{1 + \gamma(1 + \mu r)},
\end{equation}
where $r$ is the distance of the receptor from the centre of a 1D growth
cone. To determine the growth cone's estimate of the gradient, we assume
evenly spaced receptors and sample from a uniform distribution for each
location across the growth cone. Activated receptors contribute

@@

>>>>

% Note this is the SfN draft abstract, and will need rewriting.
\emph{In the classic Chemoaffinity theory, the retinotectal axon projection is
thought to use pairs of orthogonal signalling gradients in the retina to
specify the eventual location of synapses made on the surface of the
tectum/superior colliculus. Similar orthogonal gradients in the tectum provide
a coordinate system which allows the axons to match their prespecified
destination with the correct location. Although the Ephrins have been shown to
guide axons toward their destination, there has yet to emerge a complete
account of the local interactions which halt the axonal growth cones in the
correct locations to recreate the topography of the retinal cells. The model
of Simpson & Goodhill (2011) provides an account of the basic topographic
arrangement of cells on the tectum, as well as reproducing well known surgical
and genetic manipulation experiments. However, it suffers from the absence of
a local chemotactic guidance mechanism. Instead, each agent in their model is
given instantaneous knowledge of the vector that would move it toward its pre
specified destination. In addition to the globally supervised chemoaffinity
term, Simpson & Goodhill introduced a competitive interaction for space
between growth cone agents and a receptor-ligand axon-axon interaction in
order to account for the full set of experimental manipulations. Here, we
propose the replacement of the chemoaffinity term with a gradient following
model consisting of axonal growth cone agents which carry receptor molecule
expression determined by their soma's location of origin on the retina. Growth
cones move on the simulated tectum guided by two pairs of opposing, orthogonal
signalling molecules representing the Ephrin ligands. We show that with only
the chemoaffinity term and a receptor-ligand based axon-axon interaction term
(meaning that all growth cone interactions are by receptor-ligand signalling),
a full range of experimental manipulations to the retinotectal system can be
reproduced. Furthermore, we show that the observation that competition is not
and essential requirement for axons to find their way (Gosse et al., 2008) is
also accounted for by the model, due to the opposing influences of signalling
gradient pairs. Finally, we demonstrate that, assuming exponentially varying
receptor expression in the retina, ligand expression should either be
exponential if the receptor-ligand signal induces repulsion (i.e. gradient
descent) or logarithmic if the signal induces attraction (gradient
ascent). Thus, we find that a model analogous to the one we presented in James
et al (2020) that accounts for murine barrel patterning is also a candidate
mechanism for the arrangement of the more continuous retinotectal system.}

@@

Our model was inspired by the agent-based model of \cite{simpson_simple_2011}
and, in common with that work, consists of agents representing axonal growth
cones, originating from a square retina and moving upon a square tectum.
%
\citet{simpson_simple_2011} models three contributions to the movement of the growth
cones; chemoaffinity, competition for space and axon-axon interactions. The
first two of these are modelled without reference to a particular mechanism;
each agent is provided with globally-supervised information about its
instantaneous location and its target location, allowing it to move along a
vector towards the target. Competition is implemented as a movement of
constant distance along the vector between two agents if two agents are within
a threshold distance of one another. Interaction is signal-induced repulsion,
which is triggered if two agents are both within the threshold distance of
each other and have compatible EphA receptor expression.

In our model, space-based competition is omitted and chemoaffinity and
axon-axon interactions occur only via signal transmission between receptors
and ligands, which are expressed in graded patterns on both retina and
tectum. Signals induce attractive or repulsive interactions; for chemoaffinity
the interaction is to ascend or descend the gradient of ligand expression on
the tectum; for axon-axon interactions, the strength of short-range repulsion
(or attraction) is governed by the receptor-ligand expression on pairs of axons.

@@

Because there is convincing evidence that EphA and EphB receptors are
expressed in exponentially increasing
patterns \citep{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004},
we use an exponential form for retinal receptor expressions, in common with
other modelling
studies~\citep{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
%
The tectum expresses the same sets of receptors and ligands, also in
orthogonal pairs of gradients. In this work, we did not consider the effect of
reverse signalling from tectal receptors to RGC ligands so we only modelled
tectal ligand expression. Although several studies model tectal ligand
expression with exponential functions \citep{koulakov_stochastic_2004}, the
experimental evidence for ligand expression is more ambiguous (Supp. Fig
X). Correspondingly, we leave open the possibility that ligand expression may
be modelled by exponential, linear or logarithmic functions.

While we do not explicitly name $r_0$, $r_1$,
etc.~as \emph{EphA}, \emph{EphB}, the suggestion is that $r$ includes EphA,
EphB, Ryk \citep{schmitt_wntryk_2006} and
Neogenin \citep{rajagopalan_neogenin_2004} receptors and that $l$ includes the
ephrin-A, ephrin-B, Wnt3 \citep{schmitt_wntryk_2006} and
RGM \citep{monnier_rgm_2002} ligands, each of which has been shown to play a
role in retino-tectal map formation.

@@

In this model, the signal transmitted when a ligand binds to a receptor on a
growth cone can lead to one of two effects; the signal may induce an
attraction towards the ligand-expression region or a repulsion away from it.
%
For axon-tectum interactions, we implement attraction as a climbing of the
tectal ligand expression gradient and repulsion as gradient descent.
%
To determine the strength of the effect we assume a purely linear receptor
binding model, and set the movement of the growth cone to be $\sum_i \pm
r_i \times \grad l_i$.

For axon-axon interactions, attraction causes a movement along a vector
between two axons causing them to move closer together; repulsion cause the
opposite effect.

@@

%% For the noise based paper:
%% Instead, we treat the receptor binding at the growth cone with greater
%% care. Based on a Michaelis-Menten model applied to
%% receptor binding, \citet{mortimer_bayesian_2009} give an expression for the
%% probability of an individual receptor being bound as a function of the ligand
%% expression, $\gamma$, and the ligand gradient, $\mu$;
%% \begin{equation}
%% P(b=1|\gamma,\mu) = \frac{\gamma(1 + \mu r)}{1 + \gamma(1 + \mu r)},
%% \end{equation}
%% where $r$ is the distance of the receptor from the centre of a 1D growth
%% cone. To determine the growth cone's estimate of the gradient, we assume
%% evenly spaced receptors and sample from a uniform distribution for each
%% location across the growth cone. Activated receptors contribute
